---
title: "Calibrate CNN: compute a confidence score threshold for each taxon"
author: "Thelma Panaiotis"
date: "06/01/2023"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 3
    collapsed: false
    fig_width: 10
    fig_height: 8 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(reticulate)
library(grid)
library(gridExtra)
library(scales)
library(MLmetrics)
library(patchwork)

```

## Prepare data

Read predictions and scores of calibration data

```{r read_preds}
df_test <- read_csv("data/dataset/test/05.test_dataset_predictions_rf.csv", col_types = cols())  %>% select(objid, y_pred, score = score_pred)
df_true <- read_csv("data/dataset/test/04.test_dataset.csv", col_types = cols()) %>% select(objid, y_true = taxon)

df <- left_join(df_true, df_test, by = "objid") %>% select(-objid)
taxa <- sort(unique(df$y_true))
```

Within each taxon, order by score and compute cumulative error as well as rank.

```{r rank}
df <- df %>% 
  group_by(y_pred) %>% 
  arrange(score) %>% 
  mutate(
    error = y_true != y_pred,
    per_rank = row_number()/n(),
    ) %>% 
  arrange(desc(score)) %>% 
  ungroup() 
```

## Inspect data

Plot counts for true classes.

```{r count_true}
df %>% 
  count(y_true) %>% 
  arrange(desc(n)) %>% 
  mutate(y_true = fct_inorder(y_true)) %>% 
  ggplot() +
  geom_col(aes(x = y_true, y = n)) +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(title = "True classes in calibration set")
```

Plot counts for predicted classes.

```{r count_pred}
df %>% 
  count(y_pred) %>% 
  arrange(desc(n)) %>% 
  mutate(y_pred = fct_inorder(y_pred)) %>% 
  ggplot() +
  geom_col(aes(x = y_pred, y = n)) +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(title = "Predicted classes in calibration set")
```

Plot prediction score distribution.

```{r scores_dist}
df %>% 
  ggplot() +
  geom_density(aes(x = score)) +
  facet_wrap(~y_pred, scales = "free_y") +
  labs(title = "Score distribution for predicted classes")
```


## Compute cumulative precision against ordered scores

```{r cum_prec}
# Function to compute cumulative precision
cum_prec <- function(y_true, y_pred){
  res <- c()
  for (i in 1:length(y_true)){
    prec <- sum(y_true[1:i] == y_pred[1:i]) / length(y_true[1:i])
    res <- c(res, prec)
  }
  return(res)
}

calib_preds <- df %>% 
  group_by(y_pred) %>% 
  mutate(cum_prec = cum_prec(y_true, y_pred)) %>% 
  arrange(score) %>% 
  mutate(cum_prec = cummax(cum_prec)) %>% 
  ungroup()

calib_preds %>% 
  ggplot() +
  geom_path(aes(x = per_rank, y = cum_prec)) +
  facet_wrap(~y_pred)
```



## Thresholding by taxon

Make a list of taxa.

```{r taxa}
taxa <- calib_preds %>% pull(y_pred) %>% unique() %>% sort()
taxa
```

There are `r length(taxa)` taxa to compute threshold for.

Define a base error threshold and prepare a tibble to store computed thresholds. 

```{r thresholds_df}
prec_threshold_base <- 0.75

taxa_threshold <- tibble(
  taxa, 
  prec_threshold = NA,
  score_threshold = NA
  )

```

For each taxon, start with an error threshold of 75%. Plot cumulative precision VS score and VS rank and compute the score threshold associated with this precision threshold and the proportion of objects that will be kept. 


```{r}
plot_threshold <- function(calib_preds, taxon, prec_threshold = 0.8){
  #' Plot cumulative error VS score and VS rank for a given error threshold.
  #' Return the score threshold associated with this error threshold 
  #' and the proportion of objects that will be kept.  
  
  calib_taxa <- calib_preds %>% 
    filter(y_pred == taxon) %>% 
    mutate(pass = cum_prec >= prec_threshold)
  
  # Compute associated score threshold and proportion of kept data. 
  score_threshold <- calib_taxa %>% filter(pass) %>% slice(1) %>% pull(score)
  keep_threshold <- calib_taxa %>% filter(pass) %>% slice(1) %>% pull(per_rank)

  # If prec_threshold was set to 0, keep all objects
  if (prec_threshold == 0) {
    score_threshold <-  0
    keep_threshold <-  0
  }
  
  # Number of true objects for this taxon
  n <- nrow(calib_taxa)
    
  
  if (n > 20000){ # if too many objects, do not plot 
    plot <- NULL
  } else {  
    p1 <- calib_taxa %>% 
      ggplot() +
      geom_vline(aes(xintercept = score_threshold)) +
      geom_label(aes(x = score_threshold, y = 0, label = format(score_threshold, digits = 2))) +
      labs(y = "Cumulative precision")
    
    p2 <- calib_taxa %>% 
      ggplot() +
      geom_vline(aes(xintercept = keep_threshold)) + 
      geom_label(aes(x = keep_threshold, y = 0, label = percent(1-keep_threshold, digits = 2))) +
      labs(y = "Cumulative precision")
  
    if (n < 1000){ # for few objects, plot points
      p1 <- p1 + geom_point(aes(x = score, y = cum_prec, colour = pass), show.legend = F)
      p2 <- p2 + geom_point(aes(x = per_rank, y = cum_prec, colour = pass), show.legend = F)
    } else { # else, plot a path
      p1 <- p1 + geom_path(aes(x = score, y = cum_prec, colour = pass), show.legend = F)
      p2 <- p2 + geom_path(aes(x = per_rank, y = cum_prec, colour = pass), show.legend = F)
    }
  
    # Make 2 plots side by side
    plot <- grid.arrange(p1, p2, ncol=2, top = textGrob(paste0(taxon, " (n = ", n, ")"), gp=gpar(fontsize=18)))  
  }
  return(list(plot, score_threshold, keep_threshold))
}
```


### Annelida

```{r Annelida, echo=F}
taxon <- "Annelida"
index <- which(taxa == taxon)
prec_threshold <- 0.75

# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```


### Appendicularia

```{r Appendicularia, echo=F}
taxon <- "Appendicularia"
index <- which(taxa == taxon)
prec_threshold <- 0.75
  
# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```


### artefact

```{r artefact, echo=F}
taxon <- "artefact"
index <- which(taxa == taxon)
prec_threshold <- 0.75

# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```


### Cnidaria

```{r Cnidaria, echo=F}
taxon <- "Cnidaria"
index <- which(taxa == taxon)
prec_threshold <- 0.75

# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```

Discard all.

### Collodaria

```{r Collodaria, echo=F}
taxon <- "Collodaria"
index <- which(taxa == taxon)
prec_threshold <- 0.75

# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```


### Copepoda

```{r Copepoda, echo=F}
taxon <- "Copepoda"
index <- which(taxa == taxon)
prec_threshold <- 0.75

# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```


### detritus

```{r detritus, echo=F}
taxon <- "detritus"
index <- which(taxa == taxon)
prec_threshold <- 0.75

# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```


### Eumalacostraca

```{r Eumalacostraca, echo=F}
taxon <- "Eumalacostraca"
index <- which(taxa == taxon)
prec_threshold <- 0.75

# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```


### Foraminifera

```{r Foraminifera, echo=F}
taxon <- "Foraminifera"
index <- which(taxa == taxon)
prec_threshold <- 0.75

# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```

Ok for error threshold at 90%.


### Mollusca

```{r Mollusca, echo=F}
taxon <- "Mollusca"
index <- which(taxa == taxon)
prec_threshold <- 0.75

# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```


### other_living

```{r other_living, echo=F}
taxon <- "other_living"
index <- which(taxa == taxon)
prec_threshold <- 0.75

# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```


### Rhizaria

```{r Rhizaria, echo=F}
taxon <- "Rhizaria"
index <- which(taxa == taxon)
prec_threshold <- 0.75

# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```



### Salpida

```{r Salpida, echo=F}
taxon <- "Salpida"
index <- which(taxa == taxon)
prec_threshold <- 0.75

# Compute
thr <- plot_threshold(calib_preds = calib_preds, taxon = taxon, prec_threshold = prec_threshold)

# Fill the threshold dataframe
taxa_threshold$prec_threshold[index] <- prec_threshold
taxa_threshold$score_threshold[index] <- thr[[2]]
taxa_threshold$keep_threshold[index] <- thr[[3]]
```



## Thresholding summary 
```{r summary}
taxa_threshold <- taxa_threshold %>% 
  rename(taxon = taxa) %>% 
  mutate(plankton = !(taxon %in% c("artefact", "detritus"))) %>% 
  mutate(retained = 1 - keep_threshold) %>% 
  mutate(usable = retained > 0.1)

write_csv(taxa_threshold, file = "data/06.taxa_threshold.csv")

taxa_threshold %>% 
  arrange(-retained) %>% 
  mutate(taxon = factor(taxon, levels = taxon)) %>% 
  ggplot() +
  geom_col(aes(x = taxon, y = retained, fill = plankton)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(title = "Proportion of retained objects")
```


